The function-sheaf dictionary for smooth commutative group schemes over finite fields

### 2014-08-27 crelle-dictionary.tex
Added the section of T_G in the etale case, and . . .
replaced all T_G with \TrFrob{G}.

We first test the category of quasicharacter sheaves by examining its objects on \'etale commutative group schemes $G$ over $\Fq$ (\S\ref{sec:etale});
\'etale group schemes form a counterpoint to connected algebraic groups, since the component group of any smooth group scheme is an \'etale group scheme.
Our key tools for understanding the trace of Frobenius in the \'etale case are a reinterpretation of $\QC(G)$ in terms of stalks (\S\ref{ssec:stalks}) and a
relationship with the Hochschild-Serre spectral sequence (\S\ref{ssec:E}) for $\Weil{} \ltimes \bG$, where $\Weil{} \subset \Gal(\Fq/\Fq)$ is the Weil group for $\Fq$.
We define (\S\ref{ssec:S}) an isomorphism $S_G$ from $\QCiso{G}$ to the second cohomology of the total space of the spectral sequence
\[
E_2^{p,q} \ceq \Hh^p(\Weil{}, \Hh^q(\bG, \EEx)) \Rightarrow \Hh^{p+q}(\Weil{} \ltimes \bG, \EEx).
\]
Paired with the short exact sequence
\[
  0 \to \Hh^0(\Weil{},\Hh^2({\bar G},\EEx)) \to \Hh^2(E^\bullet_G) \to \Hh^1(\Weil{},\Hh^1({\bar G},\EEx)) \to 0
\]
arising from the spectral sequence, the isomorphism $S_G$ allows us to show (\S\ref{ssec:SandT}) that 
%$\TrFrob{G}$ is surjective with kernel $\Hh^2(\bG,\EEx)^{\Weil{}}$. We find that 
the group homomorphism 
\[
T_G: \QCiso{G} \to G(\Fq)^*
\]
provided by the trace of Frobenius is surjective with kernel $\Hh^2(\bG,\EEx)^{\Weil{}}$.
One can show, using the K\"unneth formula, that $\Hh^2(\bG,\EEx)$ is non-trivial unless $\bG$ is cyclic.
%The fact that $T_G$ may not be an isomorphism should not be too troubling: if quasicharacter sheaves $(\gqcs{L}, \mu,\phi)$ and $(\gqcs{L}', \mu',\phi')$ determine the same character of $G(\Fq)$ then $(\gqcs{L},\phi)$ and $(\gqcs{L}',\phi')$ are isomorphic as Weil sheaves.

Returning to the case of connected commutative algebraic groups $G$, we consider quasicharacter sheaves that arise via base change to $\bFq$ from local systems on $G$ (\S\ref{ssec:descentG}) and
those that appear in a pushforward from a constant sheaf along a discrete isogeny $H \to G$ (\S\ref{ssec:discrete_isogenies}).  While these constructions make sense even for non-connected $G$, in the connected
case we show that every quasicharacter sheaf can be described in both of these ways (\S\ref{ssec:connected}).  We use this fact to prove that $T_G$ is an isomorphism for connected commutative algebraic groups $G$
and to determine the automorphism groups of quasicharacter sheaves on such $G$.

%
%The appearance of $\Hh^2(\bG,\EEx)^{\Gal(\bFq/\Fq)}$ in these two cases is interesting: on the one hand, when $G$ is a connected commutative algebraic group over $\Fq$, isomorphism classes of quasicharacter sheaves on $G$ are measured by $\Hh^2(\bG,\EEx)^{\Gal(\bFq/\Fq)}$ and the trace of Frobenius gives an isomorphism $T_G: \QCiso{G} \to \Hom(G(\Fq),\EEx)$; on the other hand, if $G$ an etale commutative group scheme over $\Fq$, then $\Hh^2(\bG,\EEx)^{\Gal(\bFq/\Fq)}$ measures the kernel of $T_G: \QCiso{G} \to \Hom(G(\Fq),\EEx)$. 
%This observation makes plain the fact that  . . .
%Indeed, quasicharacter sheaves on \'etale commutative group schemes that are trivialized by discrete isogenies are invisible to the trace of Frobenius.

@@

The process of creating a category from the group of quasicharacters of a $p$-adic  torus informs our choice of the term quasicharacter sheaf in this paper.
We also want to situate our terminology in a historical context.
It is not uncommon to refer to local systems satisfying \eqref{introbox} on a connected, commutative algebraic group as character sheaves;
see for example, \cite{kamgarpour:09a}*{Intro}.
As we have indicated, quasicharacter sheaves evolved from this notion,
with an eye toward quasicharacters of $p$-adic groups.
However, the main use of the term character sheaf is of course due to Lusztig.
It is applied to certain perverse sheaves on connected reductive algebraic groups over algebraically closed fields in
\cite{lusztig:85a}*{Def.~2.10} and to certain perverse sheaves on reductive groups
over algebraically closed fields with finite cyclic component groups in the series of papers
beginning with \cite{lusztig:disconnected1}.
When commutative, such groups are extensions of finite cyclic groups by algebraic tori,
in which case it is not difficult to relate Frobenius-stable character sheaves to our quasicharacter sheaves.
The new features that we have found pertaining to Weil sheaves and the non-triviality of $\Hh^2(\pi_0({\bar G}),\EEx)^{\Weil{}}$ do not arise in this context.

@@

We will see that $T_G : \QCiso{G} \to G(\Fq)^*$ may not provide complete
information about isomorphism classes of quasicharacter sheaves on $G$ when $G$ is not a connected algebraic group.
Our main tool for understanding this phenomenon
is a group homomorphism $S_G: \QCiso{G} \to \Hh^2(E^\bullet_G)$ defined in Section~\ref{ssec:S}, for which the next two sections are preparation.

@@

This sequence will play a key role in understanding the kernel of $T_G$, as described in the next few sections.
For this application, we need a good understanding of these maps to and from the total complex.

@@

\begin{proposition}\label{prop:sur_etale}
If $G$ is \'etale, then $\TrFrob{G} : \QCiso{G} \to G(\Fq)^*$ is surjective.
\end{proposition}
\begin{proof}
Pick $\chi \in G(\Fq)^*$.
By Lemma~\ref{lem:dual-inv}, there is some $b\in G(\bFq)^*$ such that $b\vert_{G(\Fq)} = \chi$.
Set $\gqcs{L}_x = \EE$ for every $x\in G(\bFq)$.
Define $\mu_{x,y} : \gqcs{L}_{x+y} \to \gqcs{L}_x\otimes \gqcs{L}_y$ by $\mu_{x,y}(1) = 1 \otimes 1$ and
$\phi_{x} : \gqcs{L}_{\Frob{}(x)} \to \gqcs{L}_x$ by $\phi_{x}(1) = b(x)$.
Since $b : G(\bFq) \to \EEx$ is a group homomorphism,
condition \eqref{nohom} is satisfied with $a =1$.
So $\qcs{L} = (\gqcs{L}, \mu, \phi)$
is a quasicharacter sheaf with $\trFrob{\qcs{L}} = \chi$.
\end{proof}

@@

\begin{corollary}\label{cor:split}
\end{corollary}

>>>>

We first test the category of quasicharacter sheaves by examining its objects on \'etale commutative group schemes $G$ over $\Fq$ (\S\ref{sec:etale});
\'etale group schemes form a counterpoint to connected algebraic groups, since the component group of any smooth group scheme is an \'etale group scheme.
Our key tools for understanding the trace of Frobenius in the \'etale case are a reinterpretation of $\QC(G)$ in terms of stalks (\S\ref{ssec:stalks}) and a
relationship with the Hochschild-Serre spectral sequence (\S\ref{ssec:E}) for $\Weil{} \ltimes \bG$, where $\Weil{} \subset \Gal(\Fq/\Fq)$ is the Weil group for $\Fq$.
We define (\S\ref{ssec:S}) an isomorphism $S_G$ from $\QCiso{G}$ to the second cohomology of the total space of the spectral sequence
\[
E_2^{p,q} \ceq \Hh^p(\Weil{}, \Hh^q(\bG, \EEx)) \Rightarrow \Hh^{p+q}(\Weil{} \ltimes \bG, \EEx).
\]
Paired with the short exact sequence
\[
  0 \to \Hh^0(\Weil{},\Hh^2({\bar G},\EEx)) \to \Hh^2(E^\bullet_G) \to \Hh^1(\Weil{},\Hh^1({\bar G},\EEx)) \to 0
\]
arising from the spectral sequence, the isomorphism $S_G$ allows us to show (\S\ref{ssec:SandT}) that 
%$\TrFrob{G}$ is surjective with kernel $\Hh^2(\bG,\EEx)^{\Weil{}}$. We find that 
the group homomorphism 
\[
\TrFrob{G}: \QCiso{G} \to G(\Fq)^*
\]
provided by the trace of Frobenius is surjective with kernel $\Hh^2(\bG,\EEx)^{\Weil{}}$.
One can show, using the K\"unneth formula, that $\Hh^2(\bG,\EEx)$ is non-trivial unless $\bG$ is cyclic.
%The fact that $\TrFrob{G}$ may not be an isomorphism should not be too troubling: if quasicharacter sheaves $(\gqcs{L}, \mu,\phi)$ and $(\gqcs{L}', \mu',\phi')$ determine the same character of $G(\Fq)$ then $(\gqcs{L},\phi)$ and $(\gqcs{L}',\phi')$ are isomorphic as Weil sheaves.

Returning to the case of connected commutative algebraic groups $G$, we consider quasicharacter sheaves that arise via base change to $\bFq$ from local systems on $G$ (\S\ref{ssec:descentG}) and
those that appear in a pushforward from a constant sheaf along a discrete isogeny $H \to G$ (\S\ref{ssec:discrete_isogenies}).  While these constructions make sense even for non-connected $G$, in the connected
case we show that every quasicharacter sheaf can be described in both of these ways (\S\ref{ssec:connected}).  We use this fact to prove that $\TrFrob{G}$ is an isomorphism for connected commutative algebraic groups $G$
and to determine the automorphism groups of quasicharacter sheaves on such $G$.

%
%The appearance of $\Hh^2(\bG,\EEx)^{\Gal(\bFq/\Fq)}$ in these two cases is interesting: on the one hand, when $G$ is a connected commutative algebraic group over $\Fq$, isomorphism classes of quasicharacter sheaves on $G$ are measured by $\Hh^2(\bG,\EEx)^{\Gal(\bFq/\Fq)}$ and the trace of Frobenius gives an isomorphism $\TrFrob{G}: \QCiso{G} \to \Hom(G(\Fq),\EEx)$; on the other hand, if $G$ an etale commutative group scheme over $\Fq$, then $\Hh^2(\bG,\EEx)^{\Gal(\bFq/\Fq)}$ measures the kernel of $\TrFrob{G}: \QCiso{G} \to \Hom(G(\Fq),\EEx)$. 
%This observation makes plain the fact that  . . .
%Indeed, quasicharacter sheaves on \'etale commutative group schemes that are trivialized by discrete isogenies are invisible to the trace of Frobenius.

@@

The process of creating a category from the group of quasicharacters of a $p$-adic  torus informs our choice of the term quasicharacter sheaf in this paper.
We also want to situate our terminology in a historical context.
It is not uncommon to refer to local systems satisfying \eqref{introbox} on a connected, commutative algebraic group as character sheaves;
see for example, \cite{kamgarpour:09a}*{Intro}.
As we have indicated, quasicharacter sheaves evolved from this notion,
with an eye toward quasicharacters of $p$-adic groups.
However, the main use of the term character sheaf is of course due to Lusztig.
It is applied to certain perverse sheaves on connected reductive algebraic groups over algebraically closed fields in
\cite{lusztig:85a}*{Def.~2.10} and to certain perverse sheaves on reductive groups
over algebraically closed fields with finite cyclic component groups in the series of papers
beginning with \cite{lusztig:disconnected1}.
When commutative, such groups are extensions of finite cyclic groups by algebraic tori,
in which case it is not difficult to relate Frobenius-stable character sheaves to our quasicharacter sheaves.
The new features that we have found pertaining to Weil sheaves and the non-triviality of $\Hh^2(\pi_0({\bar G}),\EEx)^{\Weil{}}$ do not arise in that context.

@@

We will see that $\TrFrob{G} : \QCiso{G} \to G(\Fq)^*$ may not provide complete
information about isomorphism classes of quasicharacter sheaves on $G$ when $G$ is not a connected algebraic group.
Our main tool for understanding this phenomenon
is a group homomorphism $S_G: \QCiso{G} \to \Hh^2(E^\bullet_G)$ defined in Section~\ref{ssec:S}, for which the next two sections are preparation.

@@

This sequence will play a key role in understanding the kernel of $\TrFrob{G}$, as described in the next few sections.
For this application, we need a good understanding of these maps to and from the total complex.

@@

\begin{proposition}\label{prop:sur_etale}
If $G$ is \'etale, then $\TrFrob{G} : \QCiso{G} \to G(\Fq)^*$ is surjective
and admits a section.
\end{proposition}
\begin{proof}
Pick $\chi \in G(\Fq)^*$. 
Let $[\beta]\in \Hh^1(\Weil{},\bG^*)$ be the class corresponding to $\chi$ under Lemma~\ref{lem:dual-inv}.
Every representative cocycle $\beta \in Z^1(\Weil{},\bG^*)$ determines a homomorphism $\beta(\Frob{}) : G(\bFq)\to \EEx$ such that $\beta(\Frob{})\vert_{G(\Fq)} = \chi$.
Set $\gqcs{L}_x = \EE$ for every $x\in G(\bFq)$.
Define $\mu_{x,y} : \gqcs{L}_{x+y} \to \gqcs{L}_x\otimes \gqcs{L}_y$ by $\mu_{x,y}(1) = 1 \otimes 1$ and
$\phi_{x} : \gqcs{L}_{\Frob{}(x)} \to \gqcs{L}_x$ by $\phi_{x}(1) = \beta(\Frob{})(x)$.
Since $\beta(\Frob{}) : G(\bFq) \to \EEx$ is a group homomorphism,
condition \eqref{nohom} is satisfied with $a =1$.
So $\qcs{L} = (\gqcs{L}, \mu, \phi)$
is a quasicharacter sheaf with $\trFrob{\qcs{L}} = \chi$.
This shows that $\TrFrob{G}$ is surjective.
%
Now let $\beta' \in Z^1(\Weil{},\bG^*)$ be another representative for $[\beta]$
so $\beta'-\beta = d_{\Weil{}} \delta$ for some $\delta \in C^0(\Weil{},\bG^*)$ defining $d \in \Hom(G(\bFq),\EEx)$.
Let $\qcs{L}'$ be the quasicharacter sheaf on $G$ defined by $\beta'$, as above.
For each $x\in G(\bFq)$, define $\brho_x :\qcs{L}_x\to \qcs{L}'_x$ by $\brho_x(1) = d(x)$.
Since condition~\ref{qc.4} is satisfied for $\{ \brho_x \tq x\in G(\bFq)\}$, it defines a morphism $\rho : \qcs{L}\to \qcs{L}'$, which is clearly and isomorphism. 
%
We have now defined a section of $\TrFrob{G}$. 
\end{proof}

@@

Combining Propositions~\ref{prop:sur_etale} and \ref{prop:etale-iso} gives the following

\begin{corollary}\label{cor:split}
The short exact sequence
 \[
    0 \to
    \Hh^0(\Weil{},\Hh^2({\bar G},\EEx)) \to
    \Hh^2(E^\bullet_G) \to
    \Hh^1(\Weil{},\Hh^1({\bar G},\EEx)) \to
    0
 \]
 is split.
\end{corollary}
